Clone and graft:\\ Testing scientific applications as they are built

### 2015-01-15 turcksin-spring-mechanics/paper.tex
Fix a typo.
Move forward some.

So this is what we're going to consider: Imagine you are dealing with two
baseballs that
are connected by a spring. If you throw them, what are their trajectories?
Newton's law says that for each of the two balls, mass times acceleration
equals the force on the body. Let's assume for a moment that there is no air
friction and that the spring is massless, then the differential equation that
describes the motion of each of the two bodies ($i=1,2$) is
\begin{align}
  \label{eq:ode}
  m_i 
  \underbrace{\mathbf x_i''(t)}_{\text{acceleration}}
  &=
  \underbrace{m_i \mathbf g}_{\text{gravity}}
  -
  \underbrace{D (\|\mathbf x_2(t) - \mathbf x_1(t)\| - L)}_{\text{magnitude of
    spring force}}
\underbrace{\mathbf d_i}_{\text{direction of spring force}},
\\
\label{eq:ic1}
  \mathbf x_i(0) &= \mathbf x_{i,0},
\\
\label{eq:ic2}
  \mathbf x_i'(0) &= \mathbf v_{i,0},
\end{align}
where $\mathbf x_i(t)$ is the position of the $i$th body at time $t$, $\mathbf
g=(0,0,-g)^T$ is the gravity acceleration, $D$ is the spring constant of
the spring that connects the two balls, and $L$ is the rest length of the
spring. The direction vector $\mathbf d_i$ is
$\mathbf d_1 = \frac{\mathbf x_2(t) - \mathbf x_1(t)}{\|\mathbf x_2(t) - \mathbf
    x_1(t)\|}$
and
$\mathbf d_2 = \frac{\mathbf x_1(t) - \mathbf x_2(t)}{\|\mathbf x_1(t) - \mathbf
    x_2(t)\|}=-\mathbf d_1$ for the two masses.
The second and third equations are the necessary initial conditions
for this second order differential equation and denote the initial position
$\mathbf x_{i,0}$ and initial velocity $\mathbf v_{i,0}$.

@@

The program that can deal with this input looks like this:
\marginpar{After documenting spring.py, split it and simplify the first
  version to the pieces we need here}
\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]{spring.py}

When you run the program on this data, it will compute the trajectory of the
two connected balls over a number of seconds.

...maybe some output...

@@

We often have an intuitive idea of how to do the former. For example, we could
compare with another code, or we could look at special cases for which we have
an analytic solution. For our model problem, one could for example prescribe
initial velocities so that the bodies are at rest and initial positions so
that the two bodies are separate by the rest length of the spring; the two
bodies then simply fall straight down and we can easily compute on a piece of
paper where they should be at time $t$. The following input file
(\texttt{testcase-2.json}) does this:
\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-2.json}

@@

The second of the poinst above -- that testing is hard and time consuming --
is also empirically false. Once you start to do it systematically, all you
have to do is morph an existing input file into a new one that tests the
aspect you have just implemented, and add it to your testsuite. It doesn't
take very long, you just have to do it right from the start. Once you do this,
you will find how much faster it suddenly becomes to debug codes. To us, one
of the most surprising facts of developing stringent testing procedures is not
just how much better our codes become, but in fact how much safer we feel
making changes in central places of our code: without tests, we always felt
insecure and wondered what a change might do to the code, whether it is
correct in all cases, and if we might be forgetting some cases. With tests,
we find ourselves a lot more at ease: just make a change, run the 140 or so
tests and see what happens -- if it all continues to work, then that may not
be a guarantee that a change is correct, but it sure makes you sleep better at
night. It really isn't all that hard to find peace of mind!

>>>>

So this is what we're going to consider: Imagine you are dealing with two
baseballs that
are connected by a spring. If you throw them, what are their trajectories?
Newton's law says that for each of the two balls, mass times acceleration
equals the force on the body. Let's assume for a moment that there is no air
friction and that the spring is massless, then the ordinary differential
equation (ODE) that describes the motion of each of the two bodies ($i=1,2$) is
\begin{align}
  \label{eq:ode}
  m_i 
  \underbrace{\mathbf x_i''(t)}_{\text{acceleration}}
  &=
  \underbrace{m_i \mathbf g}_{\text{gravity}}
  -
  \underbrace{D (\|\mathbf x_2(t) - \mathbf x_1(t)\| - L)}_{\text{magnitude of
    spring force}}
\underbrace{\mathbf d_i}_{\text{direction of spring force}},
\\
\label{eq:ic1}
  \mathbf x_i(0) &= \mathbf x_{i,0},
\\
\label{eq:ic2}
  \mathbf x_i'(0) &= \mathbf v_{i,0},
\end{align}
where $\mathbf x_i(t)$ is the position of the $i$th body at time $t$, $\mathbf
g=(0,0,-g)^T$ is the gravity acceleration, $D$ is the spring constant of
the spring that connects the two balls, and $L$ is the rest length of the
spring. The direction vector $\mathbf d_i$ is
$\mathbf d_1 = \frac{\mathbf x_2(t) - \mathbf x_1(t)}{\|\mathbf x_2(t) - \mathbf
    x_1(t)\|}$
and
$\mathbf d_2 = \frac{\mathbf x_1(t) - \mathbf x_2(t)}{\|\mathbf x_1(t) - \mathbf
    x_2(t)\|}=-\mathbf d_1$ for the two masses.
The second and third equations are the necessary initial conditions
for this second order differential equation and denote the initial position
$\mathbf x_{i,0}$ and initial velocity $\mathbf v_{i,0}$.

@@

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
\marginpar{After documenting spring.py, split it and simplify the first
  version to the pieces we need here}
\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]{spring.py}

\begin{wrapfigure}{R}{0.2\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.18\textwidth]{tests/testcase-1.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Trajectories for the two balls with the input file from
    Section~\ref{sec:first-steps}. Note that axes are not shown at the same
    scale.}
  \vspace*{-3mm}
  \label{fig:testcase-1}
\end{wrapfigure}

When you run the program on this data, it will compute the trajectory of the
two connected balls over a number of seconds and print their final positions
to the console.
Of course, having a plot of their trajectories would be nice too, and you can
get that by adding the following lines to the bottom of the script:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
import matplotlib.pyplot
from mpl_toolkits.mplot3d import Axes3D
fig = matplotlib.pyplot.figure()
canvas = fig.gca(projection='3d')
canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2], label='body 1')
canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], label='body 2')
matplotlib.pyplot.show()
\end{lstlisting}
The output this creates is shown in Fig.~\ref{fig:testcase-1}.

@@

We often have an intuitive idea of how to do the former. For example, we could
note that the center of mass at the end time $T=5$ should have fallen a
distance of $\delta z=\frac 12 g T^2=\frac{9.81 \cdot 25}{2}\textrm m\approx
123\textrm{m}$, well in agreement with what Fig.~\ref{fig:testcase-1} shows. We
could also compare with another code. Or we could look at special cases for
which we have 
an analytic solution. For our model problem, one could for example prescribe
initial velocities so that the bodies are at rest and initial positions so
that the two bodies are separated by the rest length of the spring; the two
bodies then simply fall straight down and we can easily compute on a piece of
paper where they should be at time $t$. The following input file
(\texttt{testcase-2.json}) does this:
\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-2.json}

@@

The second of the points above -- that testing is hard and time consuming --
is also empirically false. Once you start to do it systematically, all you
have to do is morph an existing input file into a new one that tests the
aspect you have just implemented, and add it to your testsuite. It doesn't
take very long, you just have to do it right from the start. Once you do this,
you will find how much faster it suddenly becomes to debug codes. To us, one
of the most surprising facts of developing stringent testing procedures is not
just how much better our codes become, but in fact how much safer we feel
making changes in central places of our code: without tests, we always felt
insecure and wondered what a change might do to the code, whether it is
correct in all cases, and if we might be forgetting some cases. With tests,
we find ourselves a lot more at ease: just make a change, run the 140 or so
tests and see what happens -- if it all continues to work, then that may not
be a guarantee that a change is correct, but it sure makes you sleep better at
night. It really isn't all that hard to find peace of mind!
